---
title: "The Alfalfa Small-Pot Experiments: Genotype-Specific Pore-Water Geochemistry Models"
author: "Valerie Milici"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

rm(list =ls())

library(tidyverse) #clean data manipulation
library(lme4) #mixed effects models
library(lmerTest) #p-values from linear mixed models
library(MASS) #for contr.sdif
library(broom.mixed) #model diagnostics
library(performance) #model diagnostics
library(rmarkdown) #to render the document
library(knitr) # for pretty markdown tables
library(car) # for bartlett test
library(stats) #bootstrap
library(parallel) #bootstrap in parallel
library(corrr) #for PCAs
library(ggcorrplot) #PCA
library(FactoMineR) #PCA
library(factoextra) #PCA
library(ggfortify) #PCA 

dat <- read.csv("~/Desktop/Alfalfa-Terraformation-GH/Task2A/data/ModData/AllPerformanceGeochem.csv")
```

<a id="toc"></a>

# Table of Contents

 [Introduction](#item-one)
 
 [Carbon](#item-three)

<a id="item-one"></a>

# Introduction

Below are the models and results of genotype-specific pore-water geochemistry data from the July - October 2022 GCR Task 2A experiment. These models explore broad trends in how the microbes that have naturally colonized the Landscape Evolutionary Observatory (LEO) at Biosphere 2, vascular plants (Alfalfa, *Medicago sativa*), and soil moisture variation affect geochemical weathering during early pedogenesis (measured from pore-water/discharge samples). We selected 17 genotypes of alfalfa to measure the effects of intraspecific variation on the above processes. We expect that variation at the root level will have the greatest effect on pedogenesis, so we selected the genotypes based on their variation in root morphology along two axes: root length and fibrous root density. 

The pore-water samples were collected at the end of a discharge experiment that allowed us to calculate the mass balance of water in the pots. In this discharge experiment, 150 ml of DI water were added to the pots and the discharge from this final watering event provided the samples used in this analysis. Because the pots had three different baseline soil saturation levels, there is variation in the amount of discharge resulting from each pot. As a result, total discharge "Dmass_tot" is a fixed effect in all models to account for this variation. Throughout the experiment, the pots were watered such that they never produced discharge, so the pore-water/discharge samples used in these analyses represent the only discharge event. More details on the experimental design and discharge experiment are available in the detailed experimental protocol 

*These models will help us to evaluate the following questions:*

1. How does intraspecific variation affect our expectations for the effects of vascular plants on geochemical weathering rates?

2. How does intraspecific variation alter the strength of plant-microbial interactions on weathering?

3. How does abiotic variation (soil moisture) affect these relationships?  

4. To what extent can root traits explain variation in patterns and relationships, and are they a good trait to use when generating hypotheses or planning experiments?

Of the original 17 genotypes planted into this experiment, 7 genotypes had decent (50% or higher) survival rates in the LEO basalt, and for the genotype-specific models, I will filter the data to only consider these seven genotypes. The included genotypes are MOC, YON, VIR, TAS, CRI, G15, and K19. The first five genotypes (MOC - CRI) have been selected for planting on the LEO hillslopes because they produce genotype-specific patterns, G15 is an alternate option in case we fail to produce enough seed from one of the top 5 genotypes, and K19 is included so that it can be absorbed into the intercept estimate (genotype is a sum to zero contrast), allowing me to see all genotypes of interest in the model output. 

## Set the Model Contrasts

First, I will define how the model interprets soil moisture ("Level"). I am using standard difference contrasts so that the intercept can be interpreted as the "average soil moisture effect"- just like in the case of a sum-to-zero contrast- & the model output will tell us how increased soil moisture affects patterns in microbial effects on plant performance (i.e., level 2 vs level 1 & level 3 vs level 2).

Next, I am going to set the genotypes as a sum to zero contrast so that the intercept is the average genotype response and the fixed effect estimates can be interpreted as the deviance of each genotype from the average. 

```{r echo=FALSE}

outs <- c("515", "639", "A20", "D16", "R24", "SWE", "212", "105", "G74")
#genotypes that have not survived well

dat1 <- dat %>% filter(!Geno %in% outs) 

dat1$Soil <- relevel(factor(dat1$Soil),
                       ref = "C")
```


```{r Set Model Contrasts}
dat1$Level <- as.factor(dat1$Level)
contrasts(dat1$Level) <- contr.sdif(3)

dat1$Geno <- factor(dat1$Geno,
                    levels = c("CRI", "YON", "MOC",
                               "TAS", "VIR", "G15", "K19"))
contrasts(dat1$Geno) <- contr.sum(7)
```

When interpreting the model output, the following numbers corrospond to each genotype:

<a id="geno-key"></a>

1 = CRI

2 = YON

3 = MOC

4 = TAS

5 = VIR

6 = G15

<a id="item-three"></a>

## Carbon (IC & OC) 

Jump to [summary](#summary-two) of Carbon models.

### Inorganic Carbon
```{r IC Geno model selection, include=F }
ICGeno1 <- lmer(IC ~ Soil + Geno * Level + scale(Dmass_tot) +
                  (1|dis_table) + (1|table) + (1|batch_IC),
                data = dat1)

ICGeno2 <- lmer(IC ~ Soil + Level + Geno + scale(Dmass_tot) +
                  (1|dis_table) + (1|table) + (1|batch_IC),
                data = dat1)

ICGeno3 <- lmer(IC ~ Soil*Geno + Level
                + scale(Dmass_tot)  +
                 (1|dis_table) + (1|table) + (1|batch_IC),
                data = dat1)

compare_performance(ICGeno1, ICGeno2, ICGeno3, rank = T)
```

```{r include=FALSE}
anova(ICGeno3)
```

```{r IC Geno, message = F}
ICGeno <- lmer(IC ~ Soil*Geno + Level + scale(Dmass_tot)  +
                 (1|dis_table) + (1|table) + (1|batch_IC),
                data = dat1)
```

*Check the model diagnostics*

```{r IC Geno diagnostics, echo=F}

check_model(ICGeno)
```

```{r IC Geno summary table, echo=F}
ICGeno %>% tidy() %>% kable()
```

Here we see the same patterns for IC that we saw for TC. This makes sense because OC is very low in the samples, and TC is dominated by IC. 

3 - MOC has low IC while 6 - G15 has high IC. Microbes increase IC. 3 - MOC is sensitive (and positively so) with microbes and is associated with above average levels of IC in the presence of microbes. 6 - G15 is negatively associated with micrboes and has decreased IC. 

```{r}
saveRDS(ICGeno, file = "~/Desktop/Alfalfa-Terraformation-GH/Task2A/04_analyses/02_models/Output/ICGeno.RDS")
```

### Effects of biomass on inorganic carbon 
```{r IC Geno biomass, message = F}

ICGeno2 <- lmer(IC ~ Soil*Geno + Level + scale(Dmass_tot) + scale(totBM) +
                  (1|dis_table) + (1|table) + (1|batch_IC),
                data = dat1)
```

```{r IC Geno biomass diagnostics, echo=F}
check_model(ICGeno2)
```

```{r IC Geno biomass summary table, echo=F}
ICGeno2 %>% tidy() %>% kable()
```

There continues to be no relationship between biomass and IC.

### Organic Carbon

```{r OC Geno model selection, include=F }
dat$batch_NPOC <- as.factor(dat$batch_NPOC)
contrasts(dat$batch_NPOC) <- contr.sum(3)

OCGeno1 <- lmer(NPOC ~ Soil + Geno*Level + scale(Dmass_tot) + batch_NPOC +
                  (1|dis_table) + (1|table),
                data = dat1)

OCGeno2 <- lmer(NPOC ~ Soil + Geno + Level + scale(Dmass_tot) + batch_NPOC +
                  (1|dis_table) + (1|table),
                data = dat1)

OCGeno3 <- lmer(NPOC ~ Soil*Geno + Level + scale(Dmass_tot)  + batch_NPOC +
                  (1|dis_table) + (1|table),
                data = dat)

compare_performance(OCGeno1, OCGeno2, OCGeno3, rank = T)
```


```{r include=F }
anova(OCGeno2)
```


```{r OC Geno, message = F}
OCGeno <- lmer(log(NPOC) ~ Soil + Geno + Level + scale(Dmass_tot) + batch_NPOC +
                  (1|dis_table) + (1|table),
                data = filter(dat1, NPOC < 9))
```

*Check the model diagnostics*

```{r OC Geno diagnostics, echo=F}
check_model(OCGeno)
```

Eh... just ok.The model is doing a bad job with the high values. 

```{r OC Geno summary table, echo=F}
OCGeno %>% tidy() %>% kable()
```

There are quite a few genotype specific trends with OC. 2 - YON & 4 - TAS have above-average OC, while 3 - MOC, 5 - VIR, and 6 - G15 have below-average OC.  

### Effects of biomass on organic carbon 
```{r OC Geno biomass, message = F}

dat$batch_NPOC <- as.factor(dat$batch_NPOC)
contrasts(dat$batch_NPOC) <- contr.sum(3)

OCGeno2 <- lmer(log(NPOC) ~ Soil + Geno + Level + scale(Dmass_tot) + scale(totBM) + batch_NPOC +  (1|dis_table) + (1|table),
                data = filter(dat1, NPOC < 9))

```

```{r OC Geno biomass diagnostics, echo=F}
check_model(OCGeno2)
```

```{r OC Geno biomass summary table, echo=F}
OCGeno2 %>% tidy() %>% kable()
```

Once more biomass has no relationship with OC

<a id="summary-two"></a>

### Carbon Summary

There is intraspecific variation in TC, IC, and OC. The results for TC and IC are nearly identical, likely because IC dominates while there is minimal OC in the samples. 

IC is another proxy for weathering, because the weathering of basalt produces inorganic carbon molecultes (carbonates). MOC has a strong positive relationship between microbes and IC, while G15 is negatively impacted by the presence of soil microbes. 

Neither microbes nor soil moisture affects OC, but there is intraspecific variation in OC. So this is a relationship that is driven by vascular plants. Two genotypes have above average OC (YON and TAS) while two genotypes have below average OC (MOC and G15). This one is harder for me to interpret because OC is bioavailable carbon, and I would expect low values both for genotypes that produce very little OC and genotypes that quickly use OC. I wonder how the literature usually interprets variation in these values.

Across the board, biomass does not affect porewater carbon concentration.

back to [table of contents](#toc)