---
title: "Genotype-Specific Plant Performance"
author: "Valerie Milici"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

rm(list =ls())

library(tidyverse) #clean data manipulation
library(lme4) #mixed effects models
library(lmerTest) #p-values from linear mixed models
library(MASS) #for contr.sdif
library(broom.mixed) #model diagnostics
library(performance) #model diagnostics
library(rmarkdown) #to render the document
library(knitr) # for pretty markdown tables
library(stats)
library(parallel)

dat <- read.csv("~/Desktop/Alfalfa-Terraformation-GH/Task2A/data/ModData/AllPerformanceGeochem.csv")
census <- read.csv("~/Desktop/Alfalfa-Terraformation-GH/Task2A/data/ModData/FullCensusTimeSeries.csv")

outs <- c("515", "639", "A20", "D16", "R24", "SWE", "212", "105", "G74")
#genotypes that have not survived well

#Remove NAs and simplify data to one observation per plant
dat1 <- dat %>%
  #We only want to analyze Live/Sterile treatments
  filter(!is.na(Geno) &
           Soil != "P") %>%
  #Only living plants have data
  filter(status == 1) %>%
  filter(!Geno %in% outs)

census1 <- census %>% filter(!Geno %in% outs)
```

<a id="toc"></a>

# Table of Contents

 [Introduction](#item-one)
 
 [Biomass](#item-two)
 
 [Survival](#item-four)

<a id="item-one"></a>

# Introduction

Below are the models and results of genotype-specific plant performance data from the July - October 2022 GCR Task 2A experiment. These models explore genotype-specific trends in how the microbes that have naturally colonized the Landscape Evolutionary Observatory (LEO) at Biosphere 2 affect alfalfa growth, and how these trends respond to variation in soil moisture. The data have been thinned so that they only consider the genotypes that had the best survival and are the candidates for sowing onto LEO. 

*These models will help us to evaluate the following questions:*

1. How does intra-specific variation affect the relationships between plants and LEO soil microbes, and 

2. how does abiotic variation (soil moisture) affect these relationships?  

## Set the Model Contrasts

First, I will define how the model interprets soil moisture ("Level"). I am using standard difference contrasts so that the intercept can be interpreted as the "average soil moisture effect"- just like in the case of a sum-to-zero contrast- & the model output will tell us how increased soil moisture affects patterns in microbial effects on plant performance (i.e., level 2 vs level 1 & level 3 vs level 2).

I'm also changing genotype to a sum-to-zero contrast. That way the model intercept (and non-geno fixed effects) can be interpreted as the average genotypic response and the significant genotype effects will be interpreted as a genotype that deviates from the average. The genotypes will be numbered in the output; the key is as follows:

Geno1 = CRI

Geno2 = G15

Geno3 = INA

Geno4 = MOC

Geno5 = TAS

Geno6 = VIR

Geno7 = YON

Geno8 = K19 (embedded within intercept)

```{r Set Model Contrasts, results='hide'}

# Data for models of single performance metrics (biomass, photosynthesis, etc.)
dat1$Level <- as.factor(dat1$Level)
contrasts(dat1$Level) <- contr.sdif(3)

dat1$Geno <- factor(dat1$Geno,
       levels = c("CRI","G15","INA","MOC","TAS","VIR","YON", "K19"))

dat1$Geno <- as.factor(dat1$Geno)
contrasts(dat1$Geno) <- contr.sum(8)

dat1$Soil <- relevel(factor(dat1$Soil), ref = "S")

# Data for "over time" analyses (RGR and Survival)
census1$Level <- as.factor(census1$Level)
contrasts(census1$Level) <- contr.sdif(3)

census1$Geno <- factor(census1$Geno,
                       levels = c("CRI","G15","INA","MOC","TAS","VIR","YON", "K19"))
census1$Geno <- as.factor(census1$Geno)
contrasts(census1$Geno) <- contr.sum(8)

census1$Soil <- relevel(factor(census1$Soil), ref = "S")


```


<a id="item-two"></a>

# Biomass Models

Jump to [summary](#summary-one)

## Total Biomass

```{r GS Total Biomass, message = F}
BMGeno <- lmer(log(totBM) ~ Geno * Soil + Level + (1|table), 
                data = dat1)
```

*Check the model diagnostics*

```{r GS BM diagnostics, echo = F, message=F, fig.dim= c(6,8)}
check_model(BMGeno)
```

The model looks good to me. Because I've included interaction terms the VIFs for Geno and Geno:Soil are high, but that doesn't indicate an actual issue with multicollinearity in the data. Now time to review the model summary.

```{r GS BM Summary, echo = F}
BMGeno %>% tidy() %>% kable()
```
First, we see similar trends to the overall performance models. Microbes reduce plant biomass, and plants grew bigger between the low and medium soil moisture treatments, with no difference in biomass between the medium and high soil moisture treatments. 

There is variation among the genotypes. Both TAS (5) and YON (7) are slightly larger than average. 

Some genotypes show variation in how they interact with soil microbes. CRI (1) and MOC (4) were less negatively affected by soil microbes than the average genotype, while TAS (5) and YON (7) are more negatively affected by microbes than average. 

```{r GS BM Save, echo=FALSE}

saveRDS(BMGeno, file = "04_analyses/02_models/Output/BiomassGS.RDS")

```

<a id="summary-one"></a>

## Summary of Patterns in Biomass

The LEO soil microbes tend to negatively affect plant biomass, and although both roots and shoots are smaller due to microbes, the microbes seem to really reduce root growth, which results in decreased belowground biomass allocation. I think that the microbes are attacking/consuming the root tissue because this tissue would have the most bio-available nutrients in the entire pot. 

Both TAS and YON seem to be especially sensitive to soil microbes and experienced the greatest reductions in biomass when microbes are introduced, while MOC and CRI tend to be rather insensitive to microbes and are typically of similar size in both live and sterile soil. VIR is the only genotype that appears to be uniformly impacted by microbes rather than disproportionately impacted at the roots. 

[contents](#toc)

<a id="item-four"></a>

# Genotype Specific Survival Models

Jump to [summary](#summary-three)

## Survival across time
```{r GS Survival Model Data Prep, echo = F}
census3 <- census1 %>% 
  mutate(Status = ifelse(status == 1, 0, 1)) #p.mort with cloglog

census4 <- census1  %>% group_by(pot) %>% 
  arrange(CensusNo) %>%
  filter(row_number() == n()) %>%
  ungroup()
#thinned data set that contains only the final observation for each plant. 
#if censusno < 5, then the plant died before the end of the experiment.

census4$status <- ifelse(census4$CensusNo < 6, 0, census4$status)
```

## Survival to the end of the experiment

Because now we are only looking at one observation for survival (whether or not the plant was alive at the end of the experiment) we are modeling survival, not mortality. 

```{r Surv.GS model, message= F}
Surv.GS <- glmer(status ~ Geno+ Soil + Level + (1|table),
                 data = census4,
                 family = binomial(),
                 glmerControl(optimizer = "bobyqa",
                              optCtrl = list(maxfun = 10000)))
```


```{r Surv.GS Summary, echo = F}
Surv.GS %>% tidy() %>% kable()
```

Same as in the mortality model. There really isn't variation among genotypes regarding their mortality probabilities. 

<a id="summary-three"></a>

## Summary of Survival Models

The best survival models are the models that don't investigate the patterns within the genotypes. I've modeled among-genotype survival with both the top genotypes and all genotypes to see if this makes a difference and overall there is almost no variation among genotypes in there survival probabilities. The only one that stands out is that YON has a higher probability of survival throughout the duration of the experiment.

[contents](#toc)

<a id="item-six"></a>

## Summary of Genotype Specific Effects

We continue to see the general trend that performance tends to be worse in the presence of soil microbes. Some of the most interesting genotype specific results come from the biomass models where we see that TAS and YON are more sensitive to microbes, while MOC and CRI are rather robust and insensitive to microbes. 

There is very little to report for growth rates, survival, transpiration, and photosynthesis. I think my top models here will be total biomass and biomass allocation. I'll look into RGR and transpiration too, to see if those seem interesting after I simulate the difference between live and sterile to estimate the "microbe effect".

[contents](#toc)
