---
title: "The Alfalfa Small-Pot Experiments: Plant Performance Models"
author: "Valerie Milici"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)


rm(list =ls())

library(tidyverse) #clean data manipulation
library(lme4) #mixed effects models
library(lmerTest) #p-values from linear mixed models
library(MASS) #for contr.sdif
library(broom.mixed) #model diagnostics
library(performance) #model diagnostics
library(rmarkdown) #to render the document
library(knitr) # for pretty markdown tables
library(ggeffects) # for direct easy model prediction plotting
library(car) # for bartlett test

dat <- read.csv("~/Desktop/Alfalfa-Terraformation-GH/Task2A/data/
                ModData/AllPerformanceGeochem.csv") 
census <- read.csv("~/Desktop/Alfalfa-Terraformation-GH/Task2A/data/
                   ModData/FullCensusTimeSeries.csv")

#Remove NAs and simplify data to one observation per plant
BMdat <- dat %>%
  #We only want to analyze Live/Sterile treatments
  filter(!is.na(Geno) &
           Soil != "P") %>%
  #Only living plants have data
  filter(status == 1)

outs <- c("212", "SWE", "D16") #Genos with only < 3 survivors

census <- census %>% filter(!Geno %in% outs)

```

<a id="toc"></a>

# Table of Contents

 [Introduction](#item-one)
 
 [Biomass](#item-two)
 
 [Growth Rate](#item-three)
 
 [Survival](#item-four)
 
 [Plant Performance Summary](#item-seven)

<a id="item-one"></a>

# Introduction

Below are the models and results of plant performance data from the July - October 2022 GCR Task 2A experiment. These models explore broad trends in how the microbes that have naturally colonized the Landscape Evolutionary Observatory (LEO) at Biosphere 2 affect alfalfa growth, and how these trends respond to variation in soil moisture.

*These models will help us to evaluate the following questions:*

1. How do the microbes native to the LEO hillslopes affect plant performance, 

2. and how does abiotic variation (soil moisture) affect these relationships?  


*Note*: Models exploring genotype-specific trends can be found in the "Plant Performance Models: Genotype-Specific" document.

## Set the Model Contrasts

First, I will define how the model interprets soil moisture ("Level"). I am using standard difference contrasts so that the intercept can be interpreted as the "average soil moisture effect"- just like in the case of a sum-to-zero contrast- & the model output will tell us how increased soil moisture affects patterns in microbial effects on plant performance (i.e., level 2 vs level 1 & level 3 vs level 2).

```{r Set Model Contrasts}
BMdat$Level <- as.factor(BMdat$Level)
contrasts(BMdat$Level) <- contr.sdif(3)

census$Level <- as.factor(census$Level)
contrasts(census$Level) <- contr.sdif(3)
```



<a id="item-two"></a>

# Biomass Models

Jump to [summary](#summary-one)

## Total Biomass

```{r Total Biomass, message = F}
BMSoil <- lmer(log(totBM) ~ Soil * Level + (1|Geno) + (1|table), 
                data = BMdat)
```

*Check the model diagnostics*

```{r BM diagnostics, echo = F, fig.dim= c(6,8)}
check_model(BMSoil)
```

The model looks good to me. Now time to review the model summary.

```{r BM Summary, echo = F}
BMSoil %>% tidy() %>% filter(!is.na(p.value)) %>% dplyr::select(-c(1:2)) %>% kable(digits = 5)
```

We can see that plants in the sterile soil treatment had greater biomass than plants in the live soil treatment. We also see that moisture affects biomass. Regardless of soil microbe exposure, plants had greater biomass when comparing moisture treatments 1 (30-50% WHC) and 2 (55-75% WHC), but additional soil moisture did not further affect plant growth (levels 2 vs 3; 80-100% WHC).

```{r BM Save, echo=FALSE}

saveRDS(BMSoil, file = "04_analyses/02_models/Output/Biomass.RDS")

```

```{r Total Biomass plot, message=FALSE, echo=FALSE}

BMpreds <- ggemmeans(BMSoil, c("Level", "Soil"), type = "fixed")

# Create Plot

BMplot <- ggplot() +
  geom_pointrange(data = BMpreds,
                  mapping = aes(x, predicted,
                                ymin = conf.low, ymax = conf.high,
                                group = group, col = group),
                  size = 1.2,
                  position = position_dodge(width = 0.2)) +
  xlab("Percent of Water Holding Capacity") +
  ylab("Total Dry Biomass (g)") +
  scale_x_discrete(labels = c("1" = "30-50", "2" = "55-75", "3" = "80-100")) +
  scale_color_viridis_d("",
                        labels = c("L" = "Live Soil", "S" = "Sterile Soil"),
                        begin = 0.2,
                        end = 0.8) +
  theme_classic(18)

ggsave(BMplot, filename = "figures/Biomass.png",
       height = 6, width =8, units = "in")

BMplot

```
## Biomass Allocation

```{r Root Mass Fraction, message = F}

RMF <- lmer(massB/totBM ~ log(totBM) + Soil * Level + (1|Geno) + (1|table), 
                data = filter(BMdat, rootsample == "Y"))

```

The model for root mass fraction is parameterized as suggested by Poorter et al. (2012) *Frontiers in Plant Science*

*Check the model diagnostics*

```{r RMF Diagnostics, echo=F, fig.dim= c(6,8)}
check_model(RMF)
```
 


```{r RMF Summary, echo = F}
RMF %>% tidy() %>% filter(!is.na(p.value)) %>% dplyr::select(-c(1:2)) %>% kable(digits = 5)
```

As is typical (especially for alfalfa), a larger proportion of total biomass is allocated to belowground biomass than aboveground biomass. Overall, as plants become larger (total biomass increases), there is increased root allocation compared to shoot allocation. Plants in sterile soil once again differ from plants in live soil; they have greater belowground biomass allocation than do live soil plants. So, microbes affect biomass allocation patterns. Soil moisture does not affect allocation patterns in either live or sterile soil plants. 

```{r BM ratio Save, echo=FALSE}

saveRDS(RMF, file = "~/Desktop/Alfalfa-Terraformation-GH/Task2A/04_analyses/02_models/Output/RMF.RDS")

```

<a id="summary-one"></a>

## Summary of Patterns in Biomass

We can see that microbes and soil moisture influenced plant growth. In general, the soil microbes that have colonized the LEO hillslope appear to inhibit alfalfa growth, and reduce root allocation (or root prevalence as the live soil plants could have been trying to increase their roots, but the microbes may be consuming the root tissue). Soil moisture only affects plant growth between the low and intermediate moisture treatments, indicating that water is limiting in the low treatment. But, plant growth is equal between the intermediate and high treatments, indicating that plants are no longer water limited in either treatment. There is no interaction between soil moisture and the effects of microbes on plant growth, so the microbes are not responding to variation in soil moisture- only the plants. 

[contents](#toc)

<a id="item-three"></a>

# Relative Growth Rates

Jump to [summary](#summary-two)

## RGR across the entire experiment

```{r RGR entire, message = F}
RGR2 <- lmer(RGR_d ~ Soil * Level +
                  HtPrevCensus +
                  (1|Geno) + (1|table) + (1|pot) + (1|CensusNo),
                data = filter(census, CensusNo != 1 & status == 1 &
                                RGR_d < 0.03 & RGR_d > -0.02))
```

*check the model diagnostics*
```{r Entire RGR diagnostics, echo = F, fig.dim= c(6,8)}
check_model(RGR2)
```

```{r RGR2 Summary, echo = F}
RGR2 %>% tidy() %>% filter(!is.na(p.value)) %>% dplyr::select(-c(1:2)) %>% kable(digits = 5)
```

Plants in sterile soil grew faster than plants in live soil throughout the experiment. Plants grew faster with increasing soil moisture. Prior model iterations showed that an interaction between microbes and soil moisture did not improve the model, so an additive model only provides the best fit for the data.

```{r RGR Save, echo=FALSE}

saveRDS(RGR2, file = "04_analyses/02_models/Output/RGR.RDS")

```

```{r RGR Figure, message=FALSE, echo=FALSE}

rgrResult <- ggemmeans(RGR2, c("Level", "Soil"), type = "fixed")

# Create Plot

RGRplot <- ggplot() +
  geom_pointrange(data = rgrResult,
                  mapping = aes(x, predicted,
                                ymin = conf.low, ymax = conf.high,
                                group = group, col = group),
                  size = 1.2,
                  position = position_dodge(width = 0.2)) +
  xlab("Percent of Water Holding Capacity") +
  ylab("Relative Growth Rate (mm/mm/day)") +
  scale_x_discrete(labels = c("1" = "30-50", "2" = "55-75", "3" = "80-100")) +
  scale_color_viridis_d("",
                        labels = c("L" = "Live Soil", "S" = "Sterile Soil"),
                        begin = 0.2,
                        end = 0.8) +
  theme_classic(18)

ggsave(RGRplot, filename = "figures/RGR.png",
       height = 6, width =8, units = "in")

RGRplot
```

<a id="summary-two"></a>

## Summary of RGR Effects

The LEO soil microbes inhibit plant growth rates. At the very beginning of the experiment, there was an interaction between soil microbes and water availability such that at increased soil moisture growth rates were indistinguishable between plants in live and sterile soil. This suggests that plants in sterile soil were able to maximize their growth rates even in the low soil moisture treatment, but the interaction between plants and microbes caused growth rates to be water-limited. However, this interaction between soil moisture and soil microbes quickly fades, and when we consider growth rates across the entire experiment, LEO microbes inhibit plant growth rates in all water treatments. This suggests that over time, increased water resources aren't enough for plants to overcome the antagonistic relationship with soil microbes. 

[contents](#toc)

<a id="item-four"></a>

# Survival

Jump to [summary](#summary-three)

## Survival across time
```{r Survival Model Data Prep, echo = F}
census1 <- census %>% 
  mutate(Status = ifelse(status == 1, 0, 1)) #p.mort with cloglog

census2 <- census  %>% group_by(pot) %>% 
  arrange(CensusNo) %>%
  filter(row_number() == n()) %>%
  ungroup()
#thinned data set that contains only the final observation for each plant. 
#if censusno < 5, then the plant died before the end of the experiment.

census2$status <- ifelse(census2$CensusNo < 6, 0, census2$status)
```


## Survival to the end of the experiment

Because now we are only looking at one observation for survival (whether or not the plant was alive at the end of the experiment) we are modeling survival, not mortality. 

```{r Surv model, message= F}
Surv <- glmer(status ~ Soil + Level + (1|Geno) + (1|table),
                 data = census2,
                 family = binomial(),
                 glmerControl(optimizer = "bobyqa",
                              optCtrl = list(maxfun = 10000)))
```

*note:* there is no interaction between soil and level, so the best model is additive only. 

```{r Surv mod save, echo=FALSE}

saveRDS(Surv, file = "04_analyses/02_models/Output/Surv.RDS")
```


```{r Surv Summary, echo = F}
Surv %>% tidy() %>% filter(!is.na(p.value)) %>% dplyr::select(-c(1:2)) %>% kable(digits = 5)
```

Plants in sterile soil are more likely to survive to the end of the experiment than plants in live soil (77% vs 34%). Survival increases between low and intermediate soil moisture, but does not change between intermediate and high soil moisture, so plants are only water limited in a way that impacts survival at the lowest soil moisture.  

```{r Surv Figure, message=FALSE, echo = FALSE}
SurvPreds <- ggpredict(Surv, c("Level", "Soil"))

Survplot <- ggplot() +
  geom_pointrange(data = SurvPreds,
                  mapping = aes(x, predicted,
                                ymin = conf.low, ymax = conf.high,
                                group = group, col = group),
                  size = 1.2,
                  position = position_dodge(width = 0.2)) +
  xlab("Percent of Water Holding Capacity") +
  ylab("Probability of Survival") +
  scale_x_discrete(labels = c("1" = "30-50", "2" = "55-75", "3" = "80-100")) +
  scale_color_viridis_d("",
                        labels = c("L" = "Live Soil", "S" = "Sterile Soil"),
                        begin = 0.2,
                        end = 0.8) +
  theme_classic(18)

ggsave(Survplot, filename = "figures/Surv.png",
      height = 6, width =8, units = "in")

Survplot

```

<a id="summary-three"></a>

## Summary of Survival Models

When considering survival both across time and the end of the experiment, we see that soil microbes reduced plant survival. This is unsurprising given the fact that soil microbes also reduced growth rate and biomass, and small plants are much more likely to die than large plants. In general we see that plants are more likely to survive when soil moisture is at least 55% of total water holding capacity. 

Interestingly, there is a marginal relationship between microbe and moisture effects on survival rates over time. This suggests that all plants were water stressed in the low moisture treatment, and that plants exposed to soil microbes had a dual stress of water and microbes that could not be overcome by increased water availability. If we go back to the root allocation result, this could be because microbes result in decreased root allocation, so plants are less able to seek water or nutrients via root expansion. Thus, plants in live soil may be less likely to be advantaged by increased soil moisture than plants in sterile soil because of their tiny roots.

[contents](#toc)